Approximating the end of nested sampling

### 2023-08-07 paper/paper.tex
theory section of paper

\documentclass[usenatbib]{mnras}
\usepackage[T1]{fontenc}
\usepackage[pdftex]{graphicx}
\usepackage{bm}
\usepackage{savesym}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{enumitem}
\usepackage{relsize}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{layouts}
\usepackage{afterpage}
\usepackage{txfonts}
\usepackage{pgf}
\usepackage{cleveref}

\newcommand{\logL}{\{ \log \Like_i \}}
\newcommand{\thetab}{\bm{\theta}}
\newcommand{\logLm}{\log \Like_\mathrm{max}}
\newcommand{\set}[1]{\{#1\}}
\newcommand{\nlive}{n_\mathrm{live}}
\newcommand{\Like}{\mathcal{L}}
\newcommand{\DKL}{D_\mathrm{KL}}

@@

\section{Introduction}
Nested sampling is a multi-purpose algorithm invented by John Skilling which simultaneously accomplishes the two tasks of Bayesian inference, model comparison and parameter estimation \citep{skilling}. It was immediately adopted for cosmology, and is now used in a wide range of physical sciences including particle physics, materials science \citep{physical_scientists} and machine learning \citep{split}. The core algorithm is unique for being the only algorithm which estimates volumes using \textit{order statistics}, which makes high-dimensional integration feasible. It also avoids problems faced by traditional Bayesian algorithms, such as multi-modality.
\par
Many instances of software for nested sampling exist, including popular implementations such as \textsc{MultiNest} \citep{multinest} and \textsc{PolyChord} \citep{polychord} which provide flexible options for nested sampling runs, as well as post-processing packages such as \textsc{anesthetic} \citep{anesthetic} to obtain and plot sample statistics. However, currently no such software is able to estimate how long a given run should last. This is of high importance to the end user, who has no indication in the middle of a run whether it will take minutes, or weeks and months to finish; even an order of magnitude estimate is useful compared to the status quo of complete ignorance.
\par
This project sets out a novel approach for estimating the endpoint of a nested sampling run at an intermediate stage. The idea is to predict the form of the likelihood function in the region we have yet to sample from, using the information we have already gained from our previous samples. We begin in section \ref{sec:Background} with an overview of Bayesian inference and nested sampling. Section \ref{sec:Endpoint prediction} presents the theory and methodology of making endpoint predictions, before results for both toy and real examples are shown in section \ref{sec:Results}. Finally, conclusions are made in section \ref{sec:Conclusions}.

\section{Background}\label{sec:Background}
\printinunitsof{in}\prntlen{\textwidth}
We present here an overview of Bayesian inference and nested sampling; for a more detailed treatment of both subjects, we refer to \citet{sivia} as well as Skilling's original paper. 
\subsection{Bayesian inference}%
\label{sub:Bayesian inference}

Given some data $D$, Bayes' theorem tells us that the probability of some parameters $\theta$ of a model $M$ is 
\begin{equation}
	\Pr(\theta\mid D, M) = \frac{\Pr(D \mid \theta, M) \Pr(\theta \mid M)}{\Pr(D  \mid  M)}
\end{equation}
Relabelling these terms,
\begin{equation}
    P(\theta) = \frac{L(\theta) \pi(\theta)}{Z}
\end{equation}
where $\pi(\theta)$ is the \textit{prior} - what was known about a model's parameters before knowledge of the data, $P(\theta)$ is the \textit{posterior} - what is known after learning the data, and $L(\theta)$ is the \textit{likelihood} of the data given the parameters.
\par
For parameter estimation, knowledge of $L(\theta)$ and $\pi(\theta)$ is enough, but $Z$, known as the \textit{evidence}, is crucial for model comparison via Bayes factors. By definition, $Z$ is the normalisation factor 
\begin{equation}
    Z = \int L(\theta)\pi(\theta) \ d\theta
\end{equation}
which is a high-dimensional integral over the parameter space. Exponential scaling with dimension means full rasterisation is impractical for most physical applications. This motivates a sampling technique which preferentially selects for regions of high posterior mass. Traditional techniques like Markov-Chain Monte Carlo (MCMC) do not sample from the prior, so it is non-trivial to estimate the evidence from posterior samples. Nested sampling sidesteps these issues by putting the evidence first, while simultaneously yielding the posterior.

\subsection{Nested sampling}
\label{sub:Nested sampling}

We first perform a change of variable to reduce the evidence integral to a single dimension. Define the prior mass 
\begin{equation}\label{X integral}
    X(\lambda) = \int_{L(\theta)>\lambda} \pi(\theta) \ d\theta 
\end{equation}
which allows us to rewrite the evidence as 
\begin{equation}
    Z = \int_0^1 \Like(X) \ dX.
\end{equation}
The integral can be calculated by ordering the prior samples by likelihood then multiplying by the associated prior mass element. Nested sampling translates this into an algorithm which incorporates sorting \textit{de facto}, by taking a set of $\nlive$ \textit{live points} and replacing the point with the lowest likelihood with a resampled point of higher likelihood at every iteration. A choice must be made for this sampling method e.g. MCMC. The number of live points can in principle vary throughout the run, so that it depends on the iteration $k$ i.e. $n_\mathrm{live}=n_k$ \citep{dynamic_ns}. 
\par
The set of live points will therefore occupy a prior volume which tends to zero, shrinking around the maximum of the posterior. Explicit calculation of the volume $X$ within each contour is expensive, so nested sampling makes a statistical estimate using the order statistics of the remaining points: at each iteration, the volume occupied by the live points decreases by the shrinkage ratio $t$, which has probability distribution and statistics
\begin{equation}
	\Pr(t) = \nlive t^{\nlive - 1}, \quad \mathrm{E}[\log t] = -\frac{1}{\nlive}, \quad \mathrm{Var}[\log t] = \frac{1}{\nlive^2}.
\end{equation}
$X_i$ is the product of the shrinkage ratios
\begin{equation}\label{xi_formula}
    X_i = \prod_{k=0}^{i} t_k,
\end{equation}
so the $X$ themselves are also random variables. It is easier to work with $\log X$, which have the means and correlations
\begin{equation}\label{X_distribution}
	\mathrm{E}[\log X_i] = -\sum_{k=0}^{i} \frac{1}{n_k}, \quad
	\mathrm{Cov}[\log X_i, \log X_j] = \sum_{k=0}^{\mathrm{min}(i, j)} \frac{1}{n_k^2}.
\end{equation}

The evidence integral is then computed using the trapezoid rule at each step, so that
\begin{equation}
	Z = \sum_i L_i \Delta w_i = \frac{1}{2}\sum_i L_i \left( X_{i-1} - X_{i+1} \right).
\end{equation}

The algorithm terminates when the evidence in the live points falls below a user-specified fraction $\epsilon$ of the total accumulated evidence. Skilling's original paper estimates this remaining evidence $\Delta Z$ to have an upper bound of $ \Like_\mathrm{max} X_i$, while some use the average remaining likelihood. Once we have chosen to stop, the remaining evidence is included by killing the live points one by one without replacement, incrementing $Z$ as before. 
\par
The full nested sampling procedure is given in algorithm \ref{alg:ns}. 
\begin{algorithm}
\caption{Nested sampling}\label{alg:ns}
\begin{algorithmic}
\State Start with $N$ points $\theta_1, \ldots, \theta_N$ from prior
\State \hspace{\algorithmicindent} initialise $Z = 0$, $X_0 = 1$
\For{$i = 1, 2, \ldots , j$}
\State $L_i \gets \text{lowest of the current likelihood values}$
\State $X_i \gets \exp(-i/N)$
\State $w_i \gets X_{i-1} - X_i$ or $(X_{i-1} - X_{i+1})/2$
\State $Z \gets Z + L_i w_i$
\State $L_i \gets $ new point $\in \{\pi(\theta): L(\theta) > L_i\}$.
\EndFor
\State Increment $Z$ by $\langle L(\theta) \rangle X_j$.
\end{algorithmic}
\end{algorithm}

\subsection{Uncertainties and the posterior bulk}\label{subsec:Uncertainties and the posterior bulk}
\begin{figure}
\begin{center}
	\includegraphics{Figures/LX.pdf}
\end{center}
\caption{Plot showing the likelihood and the posterior weights $\Like(X)X$. The bulk of posterior mass is contained within a region of size $e^{-\DKL}$, meaning one must iterate through at least $n\DKL$ steps to fully explore the posterior.}
\label{fig:LX}
\end{figure}

\begin{figure}
\begin{center}
	\includegraphics{Figures/X_samples.pdf}
\end{center}
\caption{Plot showing samples of $\log X$ plotted against the likelihoods of the sample points. The uncertainty increases with the number of steps taken.}
\label{fig:X_samples}
\end{figure}

Most of the posterior is contained within a small fraction of the prior of size $X^{*} = e^{-\DKL}$, where
\begin{equation}
\DKL = \int P(\theta) \log  \frac{P(\theta)}{\pi(\theta)} \ d\theta
\end{equation}
is the information gain between the prior and posterior; this is visualised in figure \ref{fig:LX}. To learn anything about the posterior or calculate an accurate evidence, we must compress down to this volume. 
\par
From equation \eqref{X_distribution}, it is easy to see that for a constant number of live points $n$ $\log X$ follows Poisson statistics:
\begin{equation}
	\mathrm{E}[\log X_i] = - \frac{i}{n}, \quad \mathrm{Var}[\log X_i] = \frac{i}{n^2}.
\end{equation}
This is illustrated in figure \ref{fig:X_samples}, where samples are taken from the $X$ distribution and plotted against the likelihood values. To reach the posterior bulk we must therefore take $i^{*} = n\DKL$ steps, which translates to an uncertainty in $\log X^{*} = -\DKL \pm \sqrt{\DKL/n}$. The evidence is $Z = \langle \Like \rangle X^{*}$, or
\begin{equation}
\log Z = \langle \log \Like \rangle + \log X^{*}.
\end{equation}
The first term has low error, since the only difference that errors in $X$ makes is to slightly change the region we average over. Uncertainties in $\log Z$ are therefore dominated by the $\mathcal{O}(1/\sqrt{n})$ errors on $\log X_i$, which are much more significant than the $\mathcal{O}(n^{-2})$ quadrature errors.
\par
The next section will outline the methodology for making endpoint predictions.

\section{Endpoint prediction}\label{sec:Endpoint prediction}
We shall begin by discussing a simple ``eyeball" method for making predictions that is familiar to experienced users of nested sampling, which makes a good sanity check to compare to the more principled approach.
\subsection{Eyeball method}
At an intermediate iteration of a nested sampling run, the output might look something like that shown in the first two columns of table \ref{tab:ns_output}. 
\begin{figure}
\centering
\scalebox{0.9}{\begin{tabular}{|c|c|c|}
\hline
iteration & log Z  & $d\log Z$   \\
\hline
6000 & -34.481 & 1.317 \\ 
6200 & -32.606 & 1.875 \\ 
6400 & -30.700 & 1.906 \\ 
6600 & -29.008 & 1.692 \\ 
6800 & -27.416 & 1.591 \\ 
7000 & -25.865 & 1.551 \\ 
7200 & -24.449 & 1.416 \\ 
7400 & -23.034 & 1.415 \\ 
7600 & -21.617 & 1.417 \\ 
7800 & -20.267 & 1.349 \\ 
8000 & -18.950 & 1.317 \\
\hline
\end{tabular}}
\captionof{table}{Output of accumulated log-evidence during a nested sampling run.}
\label{tab:ns_output}
\end{figure}
\begin{figure}
  \centering
  \includegraphics{Figures/linear_extrapolate.pdf}
  \captionof{figure}{Prediction of endpoint using linear extrapolation. The estimate is much lower than the true endpoint, primarily because of the long tail in $d\log Z$.}
  \label{fig:linear_extrapolate}
\end{figure}
The outputs $\log Z$ appear to converge at a predictable rate, which is shown in terms the increments in column three. One could feasibly extrapolate these to predict when the increments approach zero, which gives an iteration number for convergence. 
\par
This is shown for a linear extrapolation in figure \ref{fig:linear_extrapolate}. The prediction is clearly an underestimate, which is mainly due to the nonlinear profile of $d\log Z$ having a long tail. However, this makes a good benchmark to beat for the more principled method using likelihood approximation.

\subsection{Likelihood approximation method}
We want to make an estimate of the number of iterations until the termination condition is met, which is when the remaining evidence falls below a fraction $\epsilon$ of the accumulated evidence.
Predictions can be made ahead of time because of the live points, which are sampled from higher likelihood regions that we have not yet compressed to. It is important to note that there will always be a small region which the live points do not penetrate because they are too diffuse, which can be seen in figure \ref{fig:dead_live_points}. 
\begin{figure}
\begin{center}
	\includegraphics{Figures/dead_live_points.pdf}
\end{center}
\caption{Plots showing the state of a nested sampling at an intermediate iteration. The live points are samples from higher likelihood regions, which allow us to see the remaining likelihood ahead of time. However, there is always a small high likelihood region where we have no samples, simply because the live points are too diffuse.}
\label{fig:dead_live_points}
\end{figure}
\par
If we could estimate the form of $\Like(X)$ at the higher likelihoods in terms of some \textit{likelihood model} $f(X, \thetab)$, we would be able to express the prior volume $X_f$ which we need to compress to as
\begin{equation}
	\Delta Z = \epsilon Z_\mathrm{tot},
\end{equation}
\begin{equation}\label{model endpoint}
	\int_0^{X_f} f(X, \thetab)\ dX = \epsilon \left( \int_0^{X_i} f(X, \thetab)\ dX + Z_\mathrm{dead} \right),
\end{equation}
where $X_i$ is the volume of the iteration we have currently compressed to, and $Z_\mathrm{dead}$ is the evidence we have accumulated up to this point. $X_f$ can then be identified by solving the above equation either analytically or numerically. Remembering that at each iteration $\log X$ decreases by $1/\nlive$, the total number of iterations $N_f$ will be
\begin{equation}
	N_f = - \nlive \log X_f .
\end{equation}

\subsubsection{Model choice}\label{sec:Model choice}
The choice of the model $f(X, \thetab)$ serves two purposes; to \textit{smooth} over the stochasticity in the live points, since $\Like(X)$ is typically a smooth function, and more importantly to \textit{extrapolate} into the region where we have no samples.
\par
Making this extrapolation necessitates assumptions about the form of the likelihood in the unsampled region. It is sensible to assume it is guided by what came before it, so we fit the model on the known likelihods of the live and dead points. Apart from this, we should make the model $f(X, \thetab)$ as flexible as possible to extrapolate without bias. We chose the model to be a Gaussian with freely varying parameters $\thetab = (\Like_\mathrm{max}, d, \sigma)$:
\begin{equation}
	f(X, \thetab) = \Like_\mathrm{max} \exp\left(-\frac{X^{2/d}}{2\sigma^2}\right).
\end{equation}
The choice was made because it is analytically tractable, and because many likelihoods in practice are near-Gaussian at their peaks. Allowing $d$ to vary freely increases flexibility, and reflects the fact that while the dimensionality of the posterior is fixed, the equivalent dimension for a Gaussian of the same information content can differ \citep{Handley_2019}.

\subsubsection{Endpoint formula}
Now we are equipped with the tools to make endpoint predictions. We will begin by fitting the model over the live points only. At each iteration, estimate the prior volumes of the live points to be those which correspond to killing off each point one by one without replacement and using formula \eqref{X_distribution}. Combined with the live likelihoods $\logL$, the parameters of the model can be found by minimising the cost function
\begin{equation}\label{chi squared}
	C^2(\thetab) = \sum_i \left| \log \Like_i - \log f(X_i, \thetab) \right| ^2
\end{equation}
with respect to $\thetab$. A key feature is that we can analytically differentiate  $C^2$ for the Gaussian model, which allows us to write the optimal values of $\logLm$ and $\sigma$ in terms of the optimal values of $d$:
\begin{equation}
    \sigma^2 = \frac{N \sum_i X_i^{4/d} - \left(\sum_i X_i^{2/d}\right)^2}{2 \sum_i \log \Like_i \sum_i X_i^{2/d} - 2N \sum_i \log \Like_i X_i^{2/d}}
\end{equation}
\begin{equation}
    \logLm = \frac{1}{N} \sum_i \logL + \frac{1}{2N\sigma^2} \sum_i X_i^{2/d} 
\end{equation}
(where $N$ here is the number of data points in we fit over), and simply minimise with respect to $d$ instead of all three variables. The result is that we are able to consistently find \textit{global} rather than local minima in a single optimisation run. We do not need to concern ourselves with the usual difficulties of global optimisation like comparing an ensemble of initial conditions, which is orders of magnitude slower.
\par
Once we have obtained the best-fit $\thetab$, the endpoint can be calculated in the manner described in \eqref{model endpoint};
\begin{equation}
	\epsilon = \frac{\int_0^{X_f} \Like_\mathrm{max} \exp\left(-X^{2/d}/2\sigma^2\right)\ dX}{\int_0^{X_i} \Like_\mathrm{max} \exp\left(-X^{2/d}/2\sigma^2\right)\ dX + Z_\mathrm{dead}}.
\end{equation}
The integrals have the analytic solution
\begin{equation}
	\int_0^{X_k} \Like_\mathrm{max} \exp\left(-X^{2/d}/2\sigma^2\right)\ dX = \frac{d}{2} \cdot 2^{d/2} \cdot \sigma^{d} \cdot \gamma_k
\end{equation}
where $\gamma_k = \gamma\left(d/2, X_k^{2/d}/2\sigma^2\right)$ is the lower incomplete gamma function. After taking the inverse of  $\gamma$ and a few more steps of algebra, we arrive at
\begin{multline}
	\log X_f = \frac{d}{2}\log 2 + d \log \sigma \\
	+ \log \gamma^{-1} \left( d/2 ,\ \epsilon \left(\gamma_i+ 2^{-d/2} \sigma^{-d} Z_\mathrm{dead}/\Like_\mathrm{max}\right)\right),
\end{multline}
and $N_f$ is just  $-\nlive$ multiplied by this. 
\par
Noting that the prediction depends on the \textit{ratio} $Z_\mathrm{dead}/\Like_\mathrm{max}$, an useful feature is that the formula is robust when we are far away from the posterior bulk. In this phase, the inferred values of $\Like_\mathrm{max}$ turn out to be very uncertain and often divergent, but the dead evidence is also negligible, so $\Like_\mathrm{max}$ has little impact on the prediction. Once $Z_\mathrm{dead}$ becomes comparable to the live evidence $\Like_\mathrm{max}$ does begin to impact the prediction, but at this point we have reached the posterior bulk, so we expect $\Like_\mathrm{max}$ to be well-estimated and thus contribute in a useful way.
\par
Note that the prediction procedure should not contribute any significant time complexity to the nested sampling run, since it involves no likelihood evaluations.

\subsubsection{Uncertainties}
Our lack of knowledge about the unsampled region means that there is always the chance that there is some huge likelihood spike at small $X$. Since there is no way to anticipate this, all uncertainty estimates are underestimates in the sense that they cannot account for this possibility; there is no estimate which will \textit{always} include the true endpoint. 
\par
We will therefore restrict ourselves to quantifying the uncertainties associated with how we have modelled the \textit{known} data, remembering that because of our lack of knowledge, the true endpoint may fall significantly outside of this range. These uncertainties have two sources: the variance in the prior volume estimates, and the uncertainty in the least squares minimisation. Since the former is the main error in nested sampling, we expect it to dominate. As we saw in section \ref{subsec:Uncertainties and the posterior bulk}, different sets of $\set{X_i}$ drawn from its probability distribution lead to different "maps" of what the remaining likelihood looks like, which lead to different endpoint predictions. The natural way to handle this is to sample from $\Pr(\bm{X})$ then find the best-fit  $\thetab$ and endpoint for each sample, from which we can obtain the mean and variance in the estimates. An visualisation of this approach is shown in figure \ref{fig:gauss_overfit}.

\begin{figure}
\begin{center}
	\includegraphics{Figures/gauss_overfit.pdf}
\end{center}
\caption{Plot showing fits of the likelihood corresponding to different draws of $\log X$. The different realisations of the likelihood lead to a set of endpoint predictions, from which a mean and variance can be obtained.}
\label{fig:gauss_overfit}
\end{figure}

\section{Results}\label{sec:Results}
\subsection{Toy data}
Let us now apply the methodology developed above in practice. We consider the following toy examples:
\begin{enumerate}[leftmargin=0.3in, label=(\alph*)]
	\item A spherically symmetric Gaussian, $\log \Like = -|\bm{r}|^2/2\sigma^2$ in $d$ dimensions
	\item A discontinous "wedding cake" likelihood with Gaussian profile and parameter $\alpha_\mathrm{w}$ controlling the plateau depth, as presented in \cite{plateaus} and shown in figure \ref{fig:wedding_3d}.
	\item A Cauchy likelihood $\log \Like = - \frac{1 + d}{2} \log (1 + |\bm{r}|^2/\gamma^2)$.
\end{enumerate}
\begin{figure}
\begin{center}
	\includegraphics{Figures/wedding_3d.pdf}
\end{center}
\caption{An example ``wedding cake" likelihood which has plateaus on top of a Gaussian profile}
\label{fig:wedding_3d}
\end{figure}

>>>>

\documentclass[usenatbib]{mnras}
\usepackage{graphicx}
\usepackage{bm}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{layouts}
\usepackage{cleveref}
\usepackage{txfonts}
\usepackage[T1]{fontenc}

\newcommand{\logL}{\{ \log \Like_i \}}
\newcommand{\thetab}{\bm{\theta}}
\newcommand{\logLm}{\log \Like_\mathrm{max}}
\newcommand{\set}[1]{\{#1\}}
\newcommand{\nlive}{n_\mathrm{live}}
\newcommand{\Like}{\mathcal{L}}
\newcommand{\DKL}{\mathcal{D}_\mathrm{KL}}

@@

\section{Introduction}
Users would like a order of magnitude runtime estimate
Figure of PolyChord output

\section{Background}
\subsection{Notation}
We shall begin with a brief description of nested sampling to establish the necessary notation. For a given likelihood $\Like(\theta)$ and prior $\pi(\theta)$, nested sampling simultaneously calculates the Bayesian evidence
\begin{equation}\label{eq:evidence}
	\mathcal{Z} = \int \Like(\theta)\ \pi(\theta)\ \mathrm{d}\theta
\end{equation}
while producing samples of the posterior distribution 
\begin{equation}
	\mathcal{P}(\theta) = \frac{\Like(\theta) \pi(\theta)}{\mathcal{Z}}.
\end{equation}
The algorithm operates by maintaining a set of $\nlive$ \textit{live points} sampled from the prior, which can vary in number throughout the run. At each iteration, the point with the lowest likelihood is removed and added to a list of \textit{dead points}. A new point is then drawn from the prior, subject to the constraint that it must have a higher likelihood than the latest dead point. Repeating the procedure leads to the live points shrinking around peaks in the likelihood. 
\par
The integral in \cref{eq:evidence} is then evaluated by transformation to a one-dimensional integral over the \textit{prior volume} $X$, defined as
\begin{equation}
	\mathcal{Z} = \int_0^1 \Like(X)\ \mathrm{d}X \approx \frac{1}{2}\sum_{i=1} \Like_i (X_{i-1}-X_{i+1}),
\end{equation}
where $X(\Like)$ is the fraction of the prior greater than $\Like$. The prior volumes $X_i$ are unknown, but can be statistically estimated as follows: one can define a \textit{shrinkage factor} $t_i$ at each iteration $X_{i} = t_i X_{i-1}$, such that
\begin{equation}
	X_i = \prod_{k=1}^i t_k.
\end{equation}
The $t_i$ are the maximum of $\nlive$ points drawn from $[0,1]$, so follow the distribution
\begin{equation}\label{eq:t_dist}
	p(t_i) = \nlive t_i^{\nlive-1}, \quad \mathrm{E}[\log t_i] = -\frac{1}{\nlive}, \quad \mathrm{Var}[\log t_i] = \frac{1}{\nlive^2}.
\end{equation}
The algorithm terminates when an user-specified condition is met; a popular choice is to terminate when the evidence in the live points falls below some fraction $\epsilon$ of the accumulated evidence e.g. $10^{-3}$. The remaining live points are then killed off one by one without replacement and added to the evidence. 
\par
Uncertainties in the evidence are dominated by the spread in the prior volume distribution, and the simplest way to estimate them is by Monte Carlo sampling over sets of $\bm{t}$.

\section{Convergence of a nested sampling run}
Let us now take a top-down view of a nested sampling run as it progresses to observe how it converges.

\subsection{Posterior convergence}
We will examine how our set of samples evolve over the course of a nested sampling run. The conventional perspective for the progress of a run is with $\log X$ along the x-axis; the prior volume is compressed at a steady rate, only picking up the bulk of the integral/posterior at around $\log X = \DKL$.  
\par
At an intermediate iteration, we have an accurate reconstruction of the posterior only down to the current prior volume $\log X_*$. The live points contain information below this volume, but not much more. If we kill off all the live points as if we were terminating, then by \cref{eq:t_dist} the prior volume of the final live point $\log X_\mathrm{min}^{\mathrm{live}}$ has mean and variance
\begin{equation}
	\mathrm{E}[\log X_\mathrm{min}^{\mathrm{live}}] = \mathrm{E}[\log X_*] - \sum_{k=1}^{\nlive} \frac{1}{n_k} \approx -\frac{i_*}{\nlive} - \log \nlive - \gamma,
\end{equation}
\begin{equation}
	\mathrm{Var}[\log X_\mathrm{min}^{\mathrm{live}}] = \mathrm{Var}[\log X_*] + \sum_{k=1}^{\nlive} \frac{1}{n_k^2} \approx \frac{i}{\nlive^2} + \frac{\pi^2}{6},
\end{equation}
where the large $\nlive$ limit is taken for the approximation and $\gamma$ is the Euler-Mascheroni constant. 
\par
This shows that the live points only get us a factor of $\log \nlive$ closer to the posterior bulk, or in other words, it is not until we are around $\log \nlive$ away from $\log X = \DKL$ that the samples begin to resemble the posterior. Intuitively, if the likelihood is sharply peaked then for most of the run the live points are too diffuse to land there with any significant probability.
\par
\begin{figure}
\begin{center}
	\includegraphics{figures/live_point_distribution.pdf}
\end{center}
\caption{}
\label{fig:}
\end{figure}

\subsection{Loglikelihood convergence}
We now offer an alternate view of the convergence of a nested sampling run, via the loglikelihood instead of the prior volumes. To get an insight into the analytics, work with a Gaussian likelihood of $d$ dimensions and a lengthscale of $\sigma$;
\begin{equation}\label{eq:logL}
	\log\Like = \log\Like_\mathrm{max} - \frac{X^{2/d}}{2\sigma^2}.
\end{equation}
The distribution of the true posterior in $\log\Like$ is
\begin{equation}
    P(\log\mathcal{L}) = \frac{1}{\Gamma(\frac{d}{2})}e^{\log\mathcal{L}-\log\mathcal{L}_\mathrm{max}} (\log\mathcal{L}_\mathrm{max}-\log\mathcal{L})^{\frac{d}{2}-1}
\end{equation}
i.e. $2(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}) \sim \chi^2_{d}$, which is the same distribution as that of the dead points up to the latest dead likelihood $\log \Like_*$. Meanwhile, the live points are uniformly distributed over the constrained prior and hence have probability distribution
\begin{equation}
	P(\log\mathcal{L}) = \frac{d}{2}\frac{(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L})^{\frac{d}{2}-1}}{(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_*)^{\frac{d}{2}}} \quad [\log\Like_* < \log\Like <\log\Like_\mathrm{max}],
    \label{eq:PL}
\end{equation}
It is helpful at this stage to define a parameter
\begin{equation}
    y = \frac{\log\mathcal{L}-\log\mathcal{L}_*}{\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_*} 
    \label{eq:y}
\end{equation}
as a normalised measure of how far a point is between the latest dead point and the maximum loglikelihood, with $y=0$ corresponding to $\mathcal{L}_*$ and $y=1$ to $\mathcal{L}_\mathrm{max}$, so that
\begin{equation}
    P(y) = \frac{d}{2}(1-y)^{\frac{d}{2}-1} \quad [0<y<1].
    \label{eq:Py}
\end{equation}
We now seek the distribution for the maximum likelihood of the live points, $\log\Like_\mathrm{max}^{\mathrm{live}}$. Using the result that the maximum of $n$ variables with cumulative distribution $F(y)$ follows $\frac{d}{dy}( 1- (1-F(y))^n)$, we obtain
\begin{equation}
    P(y_\mathrm{max}^\mathrm{live}) = \frac{nd}{2}(1-y_\mathrm{max}^\mathrm{live})^{\frac{d}{2}-1}\left(1-(1-y_\mathrm{max}^\mathrm{live})^{\frac{d}{2}}\right)^{n-1}[0<y_\mathrm{max}^\mathrm{live}<1],
    \label{eq:Pyhat}
\end{equation}
which in the limit of large live points and dimensions can be roughly summarised by
\begin{equation}\label{eq:ylivemax}
	    \lim_{d,n\to\infty} y_\mathrm{max}^\mathrm{live} \sim \frac{2\log n}{d} \pm \sqrt{\frac{2}{3}}\frac{\pi}{d}.
\end{equation}
This shows that in general the live points are nowhere near the maximum loglikelihood at any iteration, though they do steadily squeeze the interval $[\log\mathcal{L}_*,\log\mathcal{L}_\mathrm{max}]$. In particular, in high dimensions $n$ only gets us harmonically/logarithmically closer, whilst $d$ pushes us linearly further away.
\par
This remains true even at the end of the nested sampling run. To see this, we write the halting condition as:
\begin{equation}
    f = \frac{\int_0^{X_\mathrm{end}} \mathcal{L} dX}{\int_0^\infty \mathcal{L} dX}.
    \label{eq:fint}
\end{equation}
Note that we have assumed that prior effects are negligible (so $1=\infty$), and that $f \ll 1$ so that the denominator is approximately the accumulated evidence. Computing this for \cref{eq:logL} we find the answer in terms of lower incomplete gamma functions
\begin{equation}
    f = 1- \frac{\Gamma_{d/2}\left(\frac{X_\mathrm{end}^{2/d}}{2\sigma^2}\right)}{\Gamma\left(\frac{d}{2}\right)}.
    \label{eq:f}
\end{equation}
Taking the $X_\mathrm{end}\ll (\sqrt{2}\sigma)^d$ limit (almost certainly valid at termination) we find
\begin{equation}
    \lim_{X_\mathrm{end}\ll (\sqrt{2}\sigma)^d} f \approx \frac{X_\mathrm{end}}{(\sqrt{2}\sigma)^d\Gamma(1+\frac{d}{2})} = \frac{(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_\mathrm{end})^{\frac{d}{2}}}{\Gamma(1+\frac{d}{2})}.
\end{equation}
We thus have an expression relating $\mathcal{L}_\mathrm{end}$ at termination to the termination fraction $f$. This becomes yet more pleasing in the large $d$ limit, since $f^{2/d}\to 1$, we find via a Stirling approximation:
\begin{equation}
    \lim_{d\to\infty} \log\mathcal{L}_\mathrm{end} \approx \log\mathcal{L}_\mathrm{max} - \frac{d}{2e}.
\end{equation}
In the event that we keep $f$ in, we replace $\frac{d}{2e}\to \frac{d}{2e}f^{2/d}$, so we can of course battle the $\frac{d}{2e}$ term, but this becomes exponentially difficult in high dimensions.

\subsection*{Observation 3: nested sampling as a maximiser}
Putting this together, taking $\mathcal{L}_*$ in \cref{eq:y} to be $\mathcal{L}_\mathrm{end}$, and combining this with \cref{eq:ylivemax} we find
\begin{equation}
    \boxed{
        \log{\mathcal{L}}_\mathrm{max}^\mathrm{live} \approx \log\mathcal{L}_\mathrm{max} - \frac{d}{2e} + \frac{\log n}{e} \pm \frac{\pi}{\sqrt{6}e}
    },
\end{equation}
showing that in general nested sampling will finish at a contour $d/2e$ away from the maximum loglikelihood, and the final set of $n$ live points can get you $\log (n)/2e$ closer, with a chance of getting $\sim\pi/\sqrt{6}e=0.472$ closer still by statistical fluctuation.
\begin{figure*}
\begin{center}
	\includegraphics{figures/anatomy.pdf}
\end{center}
\caption{}
\label{fig:anatomy}
\end{figure*}
\par
From above analysis we learn that
\begin{itemize}
	\item Knowledge of the posterior early on in the run is limited by the maximum live point, which is far from the posterior bulk
	\item At termination, the maximum live point is far from the true maximum --> nested sampling alone is a poor optimiser (but great for finding starting point of gradient descent)
\end{itemize}

\section{Method}
Let us now tackle the problem of estimating how many steps we still have to take in a nested sampling run before we can terminate.
\par
The above analysis shows that our knowledge of the posterior at an intermediate iteration is limited by the maximum live point, which until just before the end is quite far from the posterior bulk. However, we can get an approximate image of the full posterior ahead of time, simply by \textit{extrapolating} the known likelihood profile; that is, the trajectory of $\Like(X)$ traced out by the live and dead points. 
\par
For practical purposes, one would never use this approximate posterior to do inference. However, it is more than sufficient for making a prediction for roughly how much evidence remains, and hence the point at which the run should terminate. Quantitatively, this proceeds as follows: fitting a model function $f(X, \thetab)$ to the known likelihood profile allows us to express the prior volume we need to compress to as 
\begin{equation}
	\Delta \mathcal{Z} = \epsilon \mathcal{Z}_\mathrm{tot},
\end{equation}
\begin{equation}\label{model endpoint}
	\int_0^{X_f} f(X, \thetab)\ dX = \epsilon \left( \int_0^{X_i} f(X, \thetab)\ dX + Z_\mathrm{dead} \right),
\end{equation}
where $X_i$ is the volume of the iteration we have currently compressed to, and $\mathcal{Z}_\mathrm{dead}$ is the evidence we have accumulated up to this point. $X_f$ can then be identified by solving the above equation either analytically or numerically. If $\nlive$ is constant then at each iteration $\log X$ decreases by $1/\nlive$, so the total number of iterations $N_f$ will be
\begin{equation}
	N_f = - \nlive \log X_f .
\end{equation}
One has freedom to choose the form of the fitting function $f(X, \thetab)$. Following the previous analysis, we use a Gaussian profile, which has the advantage that the regression itself can be sped up analytically, as well as the fact that many real likelihoods are Gaussian or near-Gaussian. However, any other function that is sufficiently flexible to capture the shape of the likelihood profile will do.

Fit a Gaussian profile, with advantages that a) Gaussian error common b) analytic optimisation 
Extraction of termination logX
Prediction uncertainty dominated by correlations, not least squares i.e. scatter of individual points - get by simulating X

\section{Results}
Toy examples - pure Gaussian,  wedding cake, Cauchy
Real cosmological examples - Planck

\end{document}
